---
title: Validation for Skygrid priors
---

```{r libraries, message=FALSE}
library(tidyverse)
```

# Regular Skygrid Prior

```{r message=FALSE}
data <- read_tsv("SkyGrid.log") %>% slice_tail(prop=0.9) %>%
    pivot_longer(cols=c(paste0("x.", seq(0, 4)), "y", "z"),
                 names_to="Var", values_to="Val")
```

If the regular SkyGrid prior is implemented correctly, the distribution
of the log of the first element should be normal with a mean of -2 and
a standard deviation of 0.5:

```{r }
data %>% filter(Var %in% c("x.0", "y")) %>%
    ggplot(aes(log(Val), col=Var)) +
    geom_freqpoly(aes(y=..density..), binwidth=0.1) +
    coord_cartesian(xlim=c(-4, 0))
```

(Here the variable `y` is sampled from the correct distribution for comparison.)

Similarly, the distribution of log of the 5th element should be normal with
a mean of -2 and a standard deviation of $\sqrt{(5-1)\times (1/2)^2}$:

```{r }
data %>% filter(Var %in% c("x.4", "z")) %>%
    ggplot(aes(log(Val), col=Var)) +
    geom_freqpoly(aes(y=..density..), binwidth=0.1) +
    coord_cartesian(xlim=c(-10, 5))
```

(Here the variable `z` is sampled from the correct distribution for comparison.)

# OU Skygrid Prior

If the OU SkyGrid prior is implemented correctly, all elements of the
RealParameter X should be distributed according to a LogNormal distribution
with M=-2 and S=0.5.

```{r }
data <- read_tsv("OUSkyGrid.log") %>% slice_tail(prop=0.9) %>%
    pivot_longer(cols=c(paste0("x.", seq(0, 19)), "y"), names_to="Var")

data %>% filter(Var %in% c("x.0", "x.1", "x.19", "y")) %>%
    ggplot(aes(value, col=Var)) +
    geom_freqpoly(aes(y=..density..), binwidth=0.01) +
    coord_cartesian(xlim=c(0, 0.6))
```

In the above figure, the variable `y` is sampled directly from LogNormal(-2,0.5) for
comparison.
